---
title: pmatch 0.1.3
author: Thomas Mailund
date: '2018-03-22'
slug: pmatch-0-1-3
categories:
  - Metaprogramming
  - Pattern-matching
tags:
  - pmatch
---

I have just [released version 0.1.3 of `pmatch` to CRAN](https://cran.r-project.org/web/packages/pmatch/index.html). There are not a lot of changes to this version compared to 0.1.2, but enough that I felt it would be good to get a version out before I pull the entire package apart to make it dependent on `toolbox` for its transformation function. Which it will be in 0.1.4, I think.

Version 0.1.3 have the following new features:

## Local variable bindings

Say you have a tree defined as 

```{r}
library(pmatch)
Tree := Leaf(v : numeric) | Node(left : Tree, right : Tree)
```

Then you can get components of a tree using `cases`

```{r}
tree <- Node(Leaf(1), Node(Leaf(2),Leaf(3)))
cases(tree,
      Leaf(v) -> "foo",
      Node(left, right) -> "bar")
```

but getting hold of the left and right subtrees as local variables is only easy with `cases` (or, if you are brave, you can get them as `tree$left` and `tree$right` but that is only until I change the implementation).

Now, you can assign to local variables using this notation:

```{r}
bind[left, right] <- tree
left
right
```

You can actually do more than that. The mechanism is good for returning more than one value from a function

```{r}
double <- function(x, y) list(x = 2 * x, y = 2 * y)
bind[x,y] <- double(2,3)
c(x, y)
```

and such assignments happen in parallel, so you can switch variable-values like

```{r}
bind[x,y] <- c(y,x)
c(x, y)
```

## New `cases` notation

I have written about issues with byte compilation of `cases` calls earlier. The byte compiler does not like that we assign to literals, so 

```{r}
is_leaf <- function(tree) {
   cases(tree, Leaf(v) -> TRUE, Node(left, right) -> FALSE)
}
```

would give you an error.

I have now extended the syntax so you can use `~` instead of `->`, and 

```{r}
is_leaf <- function(tree) {
   cases(tree, Leaf(v) ~ TRUE, Node(left, right) ~ FALSE)
}
```

is accepted just fine by the byte compiler (and any other static analysis programs I've tried).

## Function rewriting

I implemented this before I extended the `cases` notation, but you can also transform a function to get around the byte compiler issues.

I wrote about that [here](https://mailund.github.io/r-programmer-blog/2018/03/18/transforming-functions-with-cases-calls/) so I won't repeat it now.

